---
title: "api_dev_R"
output: html_document
---

Import the required libraries
``` {r imports}

library(anytime)
library(stringr)
library(data.table)
library(glmnet)
library(jsonlite)
library(lubridate)
library(plumber)

```

Import the airport JSON array
``` {r import JSON}

airport_list <- fromJSON("airport_encodings.json")
airports <- airport_list[,1]
```

Import the previously saved regression model
``` {r import_model}

delays_model <- readRDS("finalized_model.rda")

```

Create_airport_encoding is a function that takes two parameters:
airport: character string representing the arrival airport code
airports: matrix containing all possible arrival airports and their codes

and returns a vector consisting of all zeros except a single 1 in the position
of the selected arrival airport.  This is therefore a one-hot encoding of the
arrival airport.
``` {r one_hot_encoding}
create_airport_encoding <- function(airport, airports) {
  # create empty list the length of the airports array
  airport_encode <- vector("integer",length(airports))
  # if requested airport is in the airports array, make the value 1 at that 
  # point in the list
  found_index <- grep(airport, airports)
  if (length(found_index) != 0 ) {
    airport_encode[found_index] <- 1
    return <- airport_encode
  }
  # else return none
  else {
    return <- NULL
  }
}

```

num_seconds is a function that takes a single parameter, a time in HHMM format
as a character string, and returns a number of seconds since midnight

``` {r convert_times}

num_seconds <- function(char_time) {
  char_time <- str_pad(char_time,4,pad="0")
  atime <- as.POSIXct(char_time, tz="", format="%H%M")
  atime <- hour(atime)*3600 + minute(atime)*60 + second(atime)
  return <- atime
}

```


TODO:  write the back-end logic to provide a prediction given the inputs
The model must be passed a vector consisting of the following:
(one-hot encoded airport vector, departure time in seconds since midnight,
arrival time in seconds since midnight)
YOUR CODE GOES HERE

TODO:  write the API endpoints.  
YOUR CODE GOES HERE